TODO LIST

### 2019-03-05 LMCS/sendrev/response-letter.tex

\begin{document}

\title{Response letter to the reviews of the paper ``Descriptive Complexity for Counting Complexity Classes''}

\maketitle

We thank very much the anonymous reviewers for their useful comments. We carefully took into consideration all of them, and implemented their suggestions (in particular, all the typos mentioned in the reviews were fixed). Below please find our answers to their main comments.

### 2019-03-05 LMCS/sendrev/framework.tex
removing comments

The syntax of QSO formulae is divided in two levels. 
The first level is composed by $\so$-formulae over $\R$ (called Boolean formulae) and the second level is made by counting operators of addition and multiplication. 
For this reason, the quantifiers in $\so$ (e.g. $\exists x$ or $\exists X$) are called Boolean quantifiers and the quantifiers that make use of addition and multiplication (e.g. $\Sigma x$ or $\Pi X$) are called {\em quantitative quantifiers}.
Furthermore, $\Sigma x$ and $\Sigma X$ are called first- and second-order sum, whereas $\Pi x$ and $\Pi X$ are called first- and second-order product, respectively.
%This division between Boolean and quantitative level is essential for understanding the difference between the logic and the quantitative part. 
This separation between the Boolean and quantitative levels is essential for understanding the difference between the logic and the quantitative parts of the framework.
\martin{creo que eso quer\'ia decir esta oraci\'on}
\marcelo{OK}
Furthermore, this will later allow us to parametrize both levels of the logic in order to capture different counting complexity classes.

>>>>

The syntax of QSO formulae is divided in two levels. 
The first level is composed by $\so$-formulae over $\R$ (called Boolean formulae) and the second level is made by counting operators of addition and multiplication. 
For this reason, the quantifiers in $\so$ (e.g. $\exists x$ or $\exists X$) are called Boolean quantifiers and the quantifiers that make use of addition and multiplication (e.g. $\Sigma x$ or $\Pi X$) are called {\em quantitative quantifiers}.
Furthermore, $\Sigma x$ and $\Sigma X$ are called first- and second-order sum, whereas $\Pi x$ and $\Pi X$ are called first- and second-order product, respectively.
%This division between Boolean and quantitative level is essential for understanding the difference between the logic and the quantitative part. 
This separation between the Boolean and quantitative levels is essential for understanding the difference between the logic and the quantitative parts of the framework.
%\martin{creo que eso quer\'ia decir esta oraci\'on}
%\marcelo{OK}
Furthermore, this will later allow us to parametrize both levels of the logic in order to capture different counting complexity classes.

### 2019-03-05 LMCS/sendrev/proof-prop-rest.tex
removing comments

We divide this proof into three parts.
First, we show that $\E{1} \not\subseteq \QE{0}$. 
For this inclusion to be true, it is required to hold for an arbitrary ordered relational signature $\R$, so it suffices to show that it  is not true for at least one such a signature.
Let $\R$ be the ordered signature that contains only the relation name $<$.
Suppose that there is a $\QE{0}$ formula $\alpha$ over $\R$ which is equivalent to the $\E{1}$-formula $\sa{x} \ex{y} (x < y)$. 
That is, for every finite $\R$-structure $\A$, $\sem{\alpha}(\A) = \size{\A} - 1$.
\martin{reformul\'e por qu\'e elijo ese R}
\marcelo{OK}
Suppose that $\alpha$ is in SNF, namely, $\alpha = \sum_{i = 1}^k \alpha_i$ for some fixed $k$. 
Since $\sem{\alpha}$ is not the identically zero function, consider some $\alpha_i$ that describes a non-null function. 
Let $\alpha_i = \sa{\bar{X}}\sa{\bar{x}}\varphi(\bar{X},\bar{x})$ where $\varphi$ is quantifier-free. 
Notice that if $\length{\bar{X}} > 0$, then the function $\sem{\alpha}$ is in $\Omega(2^{\size{\A}})$, as it was proven in~\cite{SalujaST95}. 
Therefore, we have that $\alpha_i = \sa{\bar{x}}\varphi(\bar{x})$. 
We conclude our proof with the following claim, from which we conclude that $\alpha = \sum_{i = 1}^k \alpha_i$ cannot be equivalent to the $\E{1}$-formula $\sa{x} \ex{y} (x < y)$. 
\begin{clm}
	Let $\beta = \sa{\bar{x}}\varphi(\bar{x})$	where $\varphi$ is quantifier free. 
	Then the function $\sem{\beta}$ is either null, greater or equal to $n$, or is in $\Omega(n^2)$, where $n$ is the size of the input structure.
\end{clm}
\proof
Assume that $\bar x = (x_1, \ldots, x_m)$, and notice that each atomic sub-formula in $\varphi(\bar{x})$ is either $(x_i = x_j)$, $(x_i < x_j)$, $\top$ or a negation thereof, where $i,j \in \{1, \ldots, m\}$. 
	Suppose $\sem{\beta}$ is not null and consider some $\R$-structure $\A$ such that $\sem{\beta}(\A) > 0$. Hence, there exists an assignment $\bar a = (a_1, \ldots, a_m)$
%	Let $\bar{a}$ be an assignment 
	for $\bar{x}$ such that $\A\models\varphi(\bar{a})$.
	Given this assignment, define an equivalence relation $\sim$ on $\{x_1, \ldots, x_m\}$ as follows: $x_i \sim x_j$ if and only if $a_i = a_j$, and assume that $\sim$ partitions $\{x_1, \ldots, x_m\}$ into $\ell$ equivalence classes, where $\ell \geq 1$.
%	It can be seen that each assignment $\bar{a}'$ that has the same ordering over its variables\footnote{For example, if $\bar{a} = (a_1,a_2,a_3,a_4)$, such ordering may be $a_1 > a_3 = a_4 > a_2$, which has tree partitions.}	as $\bar{a}$ also satisfies $\A\models\varphi(\bar{a}')$. 
%	If this ordering has $k$ partitions then $\binom{\size{\A}}{k}$ assignments for $\bar{x}$ will satisfy it, and therefore, $\sem{\beta} \geq \binom{\size{\A}}{k}$. 
Then we have that there exist at least $\binom{\size{\A}}{\ell}$ assignments $\bar b$ for $\bar{x}$ such that $\A\models\varphi(\bar{b})$. Thus, given that if $\ell = 1$, then $\binom{\size{\A}}{\ell} = \size{\A}$, and if $\ell \geq 2$, then $\binom{\size{\A}}{\ell} \in \Omega(\size{\A}^2)$, we conclude that the claim holds.
\qed
\martin{cambie order por ordering, creo que se entiende mejor}
\marcelo{me parece que las nociones de ordering y particion no quedaban claras, por esto defini de manera precisa la nocion de particion y cambie la demostracion}
Now we show that $\QE{0} \not\subseteq \E{1}$. 
In Theorem \ref{theo-pi1-pnf} we proved that there is no formula in $\loge{1}$-PNF equivalent to the formula $\alpha = 2$. 
Every formula in $\E{1}$ can be expressed in $\loge{1}$-PNF, which implies that $2 \not\in \E{1}$. Therefore, given that $2 \in \QE{0}$ by the definition of this logic, we conclude that $\QE{0} \not\subseteq \E{1}$.

>>>>

We divide this proof into three parts.
First, we show that $\E{1} \not\subseteq \QE{0}$. 
For this inclusion to be true, it is required to hold for an arbitrary ordered relational signature $\R$, so it suffices to show that it  is not true for at least one such a signature.
Let $\R$ be the ordered signature that contains only the relation name $<$.
Suppose that there is a $\QE{0}$ formula $\alpha$ over $\R$ which is equivalent to the $\E{1}$-formula $\sa{x} \ex{y} (x < y)$. 
That is, for every finite $\R$-structure $\A$, $\sem{\alpha}(\A) = \size{\A} - 1$.
%\martin{reformul\'e por qu\'e elijo ese R}
%\marcelo{OK}

Suppose that $\alpha$ is in SNF, namely, $\alpha = \sum_{i = 1}^k \alpha_i$ for some fixed $k$. 
Since $\sem{\alpha}$ is not the identically zero function, consider some $\alpha_i$ that describes a non-null function. 
Let $\alpha_i = \sa{\bar{X}}\sa{\bar{x}}\varphi(\bar{X},\bar{x})$ where $\varphi$ is quantifier-free. 
Notice that if $\length{\bar{X}} > 0$, then the function $\sem{\alpha}$ is in $\Omega(2^{\size{\A}})$, as it was proven in~\cite{SalujaST95}. 
Therefore, we have that $\alpha_i = \sa{\bar{x}}\varphi(\bar{x})$. 
We conclude our proof with the following claim, from which we conclude that $\alpha = \sum_{i = 1}^k \alpha_i$ cannot be equivalent to the $\E{1}$-formula $\sa{x} \ex{y} (x < y)$. 
\begin{clm}
	Let $\beta = \sa{\bar{x}}\varphi(\bar{x})$	where $\varphi$ is quantifier free. 
	Then the function $\sem{\beta}$ is either null, greater or equal to $n$, or is in $\Omega(n^2)$, where $n$ is the size of the input structure.
\end{clm}
\proof
Assume that $\bar x = (x_1, \ldots, x_m)$, and notice that each atomic sub-formula in $\varphi(\bar{x})$ is either $(x_i = x_j)$, $(x_i < x_j)$, $\top$ or a negation thereof, where $i,j \in \{1, \ldots, m\}$. 
	Suppose $\sem{\beta}$ is not null and consider some $\R$-structure $\A$ such that $\sem{\beta}(\A) > 0$. Hence, there exists an assignment $\bar a = (a_1, \ldots, a_m)$
%	Let $\bar{a}$ be an assignment 
	for $\bar{x}$ such that $\A\models\varphi(\bar{a})$.
	Given this assignment, define an equivalence relation $\sim$ on $\{x_1, \ldots, x_m\}$ as follows: $x_i \sim x_j$ if and only if $a_i = a_j$, and assume that $\sim$ partitions $\{x_1, \ldots, x_m\}$ into $\ell$ equivalence classes, where $\ell \geq 1$.
%	It can be seen that each assignment $\bar{a}'$ that has the same ordering over its variables\footnote{For example, if $\bar{a} = (a_1,a_2,a_3,a_4)$, such ordering may be $a_1 > a_3 = a_4 > a_2$, which has tree partitions.}	as $\bar{a}$ also satisfies $\A\models\varphi(\bar{a}')$. 
%	If this ordering has $k$ partitions then $\binom{\size{\A}}{k}$ assignments for $\bar{x}$ will satisfy it, and therefore, $\sem{\beta} \geq \binom{\size{\A}}{k}$. 
Then we have that there exist at least $\binom{\size{\A}}{\ell}$ assignments $\bar b$ for $\bar{x}$ such that $\A\models\varphi(\bar{b})$. Thus, given that if $\ell = 1$, then $\binom{\size{\A}}{\ell} = \size{\A}$, and if $\ell \geq 2$, then $\binom{\size{\A}}{\ell} \in \Omega(\size{\A}^2)$, we conclude that the claim holds.
\qed
%\martin{cambie order por ordering, creo que se entiende mejor}
%\marcelo{me parece que las nociones de ordering y particion no quedaban claras, por esto defini de manera precisa la nocion de particion y cambie la demostracion}

Now we show that $\QE{0} \not\subseteq \E{1}$. 
In Theorem \ref{theo-pi1-pnf} we proved that there is no formula in $\loge{1}$-PNF equivalent to the formula $\alpha = 2$. 
Every formula in $\E{1}$ can be expressed in $\loge{1}$-PNF, which implies that $2 \not\in \E{1}$. Therefore, given that $2 \in \QE{0}$ by the definition of this logic, we conclude that $\QE{0} \not\subseteq \E{1}$.

### 2019-03-05 LMCS/sendrev/proof-sigma2hard.tex
removing comments

Next, we prove that $\shdhsat$ is hard for $\eqso(\ehorn)$ over each signature~$\R$ under parsimonious reductions. For each $\eqso(\ehorn)$ formula $\alpha$ over $\R$, we define a polynomial-time function $g_{\alpha}$ that receives an $\R$-structure $\A$ and outputs an instance of $\shdhsat$ such that $\sem{\alpha}(\A) = \shdhsat(g_{\alpha}(\A))$. By Proposition~\ref{theo-pnf-snf}, we can assume that $\alpha$ is of the form:
$$
\alpha \ = \ \sum_{i = 1}^{m} \sa{\bar{X}_{i}}\sa{\bar{x}} \ex{\bar{y}} \bigwedge_{j = 1}^{n} \fa{\bar{z}} \varphi^i_j(\bar{X}_{i},\bar{x},\bar{y},\bar{z}),
$$
where each $\varphi^i_j$ is a Horn clause, and each $\bar{X}_{i}$ is a sequence of second-order variables.
Consider $\bar{X}$ as the union of all $\bar{X}_{i}$. We replace each of the $m$ summands in $\alpha$ with
$$
\sa{\bar{X}}\sa{\bar{x}} \ex{\bar{y}} \bigg( \bigwedge_{j = 1}^{n} \fa{\bar{z}} \varphi^i_j(\bar{X}_{i},\bar{x},\bar{y},\bar{z})\wedge\bigwedge_{X\not\in\bar{X}_{i}}\!\!\fa{\bar{u}}X(\bar{u})\bigg),
$$
whose sum is equivalent to $\alpha$.
Now, consider a finite $\R$-structure $\A$ with domain $A$. 
The next transformation of $\alpha$ and $\A$ towards a disjunction of Horn-formulae is to expand each first-order quantifier (i.e. $\Sigma{\bar{x}}$,  $\exists\bar{y}$, and $\forall\bar{z}$) by replacing variables with constants.
%their variables with first-order constants. 
More specifically, we obtain the following formula that defines the same function as $\alpha$, and it is of polynomial size in the size of $\A$ (recall that $\alpha$ is fixed):
$$
\alpha_{\A} \ = \ \sum_{i = 1}^{m} \sum_{\bar{a}\,\in A^{\length{\bar{x}}}} \sa{\bar{X}}\bigvee_{\bar{b}\,\in A^{\length{\bar{y}}}}\bigg(\bigwedge_{j = 1}^n\bigwedge_{\bar{c}\,\in A^{\length{\bar{z}}}}\varphi^i_j(\bar{X}_i,\bar{a},\bar{b},\bar{c})\wedge \bigwedge_{X \not\in \bar{X}_i}\bigwedge_{\bar{e}\,\in A^{\arity(X)}}\!\!\!\!\!\!\! X(\bar{e})\bigg).
$$
\martin{reescribi esta formula}
\marcelo{OK. Le agregue los parentesis grandes a esta formula y la anterior para que quedara mas claro el scope del cuantificador existencial y la disyuncion}
Notice that each first-order subformula in $\varphi^i_j(\bar{X}_i,\bar{a},\bar{b},\bar{c})$ has no free variables and, therefore, we can evaluate each of them in polynomial time and easily rewrite $\alpha_{\A}$ to an equivalent formula that does not have any first-order subformula. In other words, in polynomial time we can replace $\varphi^i_j$ with a disjunction of negative literals of the form $\neg X_{\ell}(\bar d)$ and at most one positive literal of the form $X_{\ell}(\bar d)$, where $\bar d$ is a tuple of constants. After simplifying, grouping and reordering the previous formula, we can obtain an equivalent formula $\alpha_{\A}'$ of the form:
$$
\alpha_{\A}' \ = \ \sum_{i = 1}^{m'}\sa{\bar{X}}\bigvee_{j = 1}^{n_1'}\bigwedge_{k = 1}^{n_2'}\psi^{i}_{j,k}(\bar{X}),
$$
where every $\psi^{i}_{j,k}(\bar{X})$ is a disjunction of 
%$\neg X_{\ell}$ and at most one $X_{\ell}$.
negative literals of the form $\neg X_{\ell}(\bar d)$ and at most one positive literal of the form $X_{\ell}(\bar d)$, where $\bar d$ is a tuple of constants. 
%It is straightforward to check that~$\sem{\alpha}(\A) = \sem{\alpha_{\A}'}(\A)$.

@@

\marcelo{Hice algunos cambios en esta demostracion (teorema \ref{sigma2hard}), por favor revisen con cuidado si todo esta bien}

>>>>

Next, we prove that $\shdhsat$ is hard for $\eqso(\ehorn)$ over each signature~$\R$ under parsimonious reductions. For each $\eqso(\ehorn)$ formula $\alpha$ over $\R$, we define a polynomial-time function $g_{\alpha}$ that receives an $\R$-structure $\A$ and outputs an instance of $\shdhsat$ such that $\sem{\alpha}(\A) = \shdhsat(g_{\alpha}(\A))$. By Proposition~\ref{theo-pnf-snf}, we can assume that $\alpha$ is of the form:
$$
\alpha \ = \ \sum_{i = 1}^{m} \sa{\bar{X}_{i}}\sa{\bar{x}} \ex{\bar{y}} \bigwedge_{j = 1}^{n} \fa{\bar{z}} \varphi^i_j(\bar{X}_{i},\bar{x},\bar{y},\bar{z}),
$$
where each $\varphi^i_j$ is a Horn clause, and each $\bar{X}_{i}$ is a sequence of second-order variables.
Consider $\bar{X}$ as the union of all $\bar{X}_{i}$. We replace each of the $m$ summands in $\alpha$ with
$$
\sa{\bar{X}}\sa{\bar{x}} \ex{\bar{y}} \bigg( \bigwedge_{j = 1}^{n} \fa{\bar{z}} \varphi^i_j(\bar{X}_{i},\bar{x},\bar{y},\bar{z})\wedge\bigwedge_{X\not\in\bar{X}_{i}}\!\!\fa{\bar{u}}X(\bar{u})\bigg),
$$
whose sum is equivalent to $\alpha$.
Now, consider a finite $\R$-structure $\A$ with domain $A$. 
The next transformation of $\alpha$ and $\A$ towards a disjunction of Horn-formulae is to expand each first-order quantifier (i.e. $\Sigma{\bar{x}}$,  $\exists\bar{y}$, and $\forall\bar{z}$) by replacing variables with constants.
%their variables with first-order constants. 
More specifically, we obtain the following formula that defines the same function as $\alpha$, and it is of polynomial size in the size of $\A$ (recall that $\alpha$ is fixed):
$$
\alpha_{\A} \ = \ \sum_{i = 1}^{m} \sum_{\bar{a}\,\in A^{\length{\bar{x}}}} \sa{\bar{X}}\bigvee_{\bar{b}\,\in A^{\length{\bar{y}}}}\bigg(\bigwedge_{j = 1}^n\bigwedge_{\bar{c}\,\in A^{\length{\bar{z}}}}\varphi^i_j(\bar{X}_i,\bar{a},\bar{b},\bar{c})\wedge \bigwedge_{X \not\in \bar{X}_i}\bigwedge_{\bar{e}\,\in A^{\arity(X)}}\!\!\!\!\!\!\! X(\bar{e})\bigg).
$$
%\martin{reescribi esta formula}
%\marcelo{OK. Le agregue los parentesis grandes a esta formula y la anterior para que quedara mas claro el scope del cuantificador existencial y la disyuncion}

Notice that each first-order subformula in $\varphi^i_j(\bar{X}_i,\bar{a},\bar{b},\bar{c})$ has no free variables and, therefore, we can evaluate each of them in polynomial time and easily rewrite $\alpha_{\A}$ to an equivalent formula that does not have any first-order subformula. In other words, in polynomial time we can replace $\varphi^i_j$ with a disjunction of negative literals of the form $\neg X_{\ell}(\bar d)$ and at most one positive literal of the form $X_{\ell}(\bar d)$, where $\bar d$ is a tuple of constants. After simplifying, grouping and reordering the previous formula, we can obtain an equivalent formula $\alpha_{\A}'$ of the form:
$$
\alpha_{\A}' \ = \ \sum_{i = 1}^{m'}\sa{\bar{X}}\bigvee_{j = 1}^{n_1'}\bigwedge_{k = 1}^{n_2'}\psi^{i}_{j,k}(\bar{X}),
$$
where every $\psi^{i}_{j,k}(\bar{X})$ is a disjunction of 
%$\neg X_{\ell}$ and at most one $X_{\ell}$.
negative literals of the form $\neg X_{\ell}(\bar d)$ and at most one positive literal of the form $X_{\ell}(\bar d)$, where $\bar d$ is a tuple of constants. 
%It is straightforward to check that~$\sem{\alpha}(\A) = \sem{\alpha_{\A}'}(\A)$.

@@

%\marcelo{Hice algunos cambios en esta demostracion (teorema \ref{sigma2hard}), por favor revisen con cuidado si todo esta bien}

### 2019-03-05 LMCS/sendrev/proof-sigmafo-minusone.tex
removing comments

For the sake of readability, we divide the proof into three parts. The last part, i.e. subtraction by one, is the most technical proof in the paper. Since this proof is several pages long, a line was drawn on the left margin to visually differentiate it from the rest of the paper.
\martin{Ac\'a menciono la l\'inea del margen izquierdo}
\marcelo{OK}

@@

Now, we define a series of rewrites of $\varphi$ that will make each formula $\varphi_i$ satisfy the following three conditions: (a) no variable from $\bar{x}$ appears in $\varphi_i^{-}(\bar{X},\bar{x},\bar{y})\wedge\varphi_i^{+}(\bar{X},\bar{x},\bar{y})$, (b) $\varphi_i^{\fo}(\bar{x},\bar{y})$ defines a weak ordering over the variables in $(\bar{x},\bar{y})$ (the precise definition of a weak ordering is detailed below) and (c) if both $X_j(\bar{z})$ and $\neg X_j(\bar{w})$ appear in the formula, the weak ordering should not satisfy $\bar{z} = \bar{w}$.
When these conditions are met, $\varphi_i'(\bar{X}, \bar{x}, \bar{y})$ will be satisfiable if and only if $\varphi_i^{\fo}(\bar{x},\bar{y})$ is satisfiable. This is proven in Claim~\ref{claim:minusone}.
We explain below how to rewrite $\varphi_i$ in order to satisfy each condition.
\martin{Peque\~na explicaci\'on de para qu\'e sirven las condiciones a b y c}
\marcelo{OK}

@@

Now, we have all the ingredients in order to define $\kappa(\alpha)$. 
Intuitively, we want to exclude the assignment $(V^*, v^*)$ from the satisfying assignments of $\varphi_i(\bar{X},\bar{x})$.
Towards this goal, we can define a formula $\psi_i(\bar{X},\bar{x})$ such that, if $\varphi_i(\bar{X}, \bar{x})$ is satisfiable, $(\A, V, v) \models \psi_i(\bar{X},\bar{x})$ if, and only if either $V \neq V^*$ or $v \neq v^*$. 
This property can be defined as follows:
\begin{align}
\psi_i(\bar{X},\bar{x}) \; := \;  &\big( \, \ex{\bar{x}'} \ex{\bar{y}} \varphi^{\fo}_i(\bar{x}',\bar{y}) \,\big) \rightarrow  \label{eq:part1} \\
&\Big( 
\ex{\bar{y}}\!\big[ \text{{\it min}-}\varphi^{\fo}_i(\bar{x}, \bar{y}) \wedge \big( \, \varphi_i'(\bar{X},\bar{x}, \bar{y})  \rightarrow \!\! \bigvee_{X \in \bar{X}}\! \ex{\bar{z}}\!( \, X(\bar{z}) \wedge   \!\!\!\!\!\!\!\!\!\!\!\! \bigwedge\limits_{X(\bar{w}) \, \in \, \varphi^{+}_i(\bar{X},\bar{y})} \!\!\!\!\!\!\!\!\!\!\!\! \bar{w}\neq\bar{z} \, ) \, \big)\big] \;  \vee \!  \label{eq:part2} \\ 
&\; \ex{\bar{x}'} \ex{\bar{y}}\big( \varphi_i'(\bar{X},\bar{x}',\bar{y})\wedge \bar{x}' < \bar{x}\big) \; \;\; \Big)  \label{eq:part3}
\end{align}
\martin{agregu\'e par\'entesis en varias f\'ormulas en esta p\'agina y las anteriores}
\marcelo{OK}
To understand the formula, first notice that the premise of the implication at (\ref{eq:part1}) is true if, and only if, $\varphi_i(\bar{X}, \bar{x})$ is satisfiable. 
Indeed, by Claim~\ref{claim:minusone} we know that if $\ex{\bar{x}} \ex{\bar{y}} \varphi^{\fo}_i(\bar{x},\bar{y})$ is true, then there exist assignments $V$ and $v$ such that $(\A,V,v) \models\varphi_i'(\bar{X},\bar{x},\bar{y})$.
The conclusion of the implication (divided into (\ref{eq:part2}) and (\ref{eq:part3})), takes care that $V(\bar{X}) \neq V^*(\bar{X})$ or $v(\bar{x}) \neq v^*(\bar{x})$.
Here, the first disjunct (\ref{eq:part2}) checks that $V(\bar{X}) \neq V^*(\bar{X})$ by defining that if $\varphi_i'(\bar{X},\bar{x}, \bar{y})$ is satisfied then $V^*(\bar{X}) \subsetneq V(\bar{X})$. 
The second disjunct (\ref{eq:part3}) is satisfied when $v(\bar{x})$ is not the lexicographically smallest tuple that satisfies $\varphi_i$ (i.e. $v(\bar{x}) \neq v^*(\bar{x})$).
Finally, from the previous discussion one can easily check that $\psi_i(\bar{X},\bar{x})$ satisfies the desired property.

>>>>

For the sake of readability, we divide the proof into three parts. The last part, i.e. subtraction by one, is the most technical proof in the paper. Since this proof is several pages long, a line was drawn on the left margin to visually differentiate it from the rest of the paper.
\medskip

@@

Now, we define a series of rewrites of $\varphi$ that will make each formula $\varphi_i$ satisfy the following three conditions: (a) no variable from $\bar{x}$ appears in $\varphi_i^{-}(\bar{X},\bar{x},\bar{y})\wedge\varphi_i^{+}(\bar{X},\bar{x},\bar{y})$, (b) $\varphi_i^{\fo}(\bar{x},\bar{y})$ defines a weak ordering over the variables in $(\bar{x},\bar{y})$ (the precise definition of a weak ordering is detailed below) and (c) if both $X_j(\bar{z})$ and $\neg X_j(\bar{w})$ appear in the formula, the weak ordering should not satisfy $\bar{z} = \bar{w}$.
When these conditions are met, $\varphi_i'(\bar{X}, \bar{x}, \bar{y})$ will be satisfiable if and only if $\varphi_i^{\fo}(\bar{x},\bar{y})$ is satisfiable. This is proven in Claim~\ref{claim:minusone}.
We explain below how to rewrite $\varphi_i$ in order to satisfy each condition.
%\martin{Peque\~na explicaci\'on de para qu\'e sirven las condiciones a b y c}
%\marcelo{OK}

@@

Now, we have all the ingredients in order to define $\kappa(\alpha)$. 
Intuitively, we want to exclude the assignment $(V^*, v^*)$ from the satisfying assignments of $\varphi_i(\bar{X},\bar{x})$.
Towards this goal, we can define a formula $\psi_i(\bar{X},\bar{x})$ such that, if $\varphi_i(\bar{X}, \bar{x})$ is satisfiable, $(\A, V, v) \models \psi_i(\bar{X},\bar{x})$ if, and only if either $V \neq V^*$ or $v \neq v^*$. 
This property can be defined as follows:
\begin{align}
\psi_i(\bar{X},\bar{x}) \; := \;  &\big( \, \ex{\bar{x}'} \ex{\bar{y}} \varphi^{\fo}_i(\bar{x}',\bar{y}) \,\big) \rightarrow  \label{eq:part1} \\
&\Big( 
\ex{\bar{y}}\!\big[ \text{{\it min}-}\varphi^{\fo}_i(\bar{x}, \bar{y}) \wedge \big( \, \varphi_i'(\bar{X},\bar{x}, \bar{y})  \rightarrow \!\! \bigvee_{X \in \bar{X}}\! \ex{\bar{z}}\!( \, X(\bar{z}) \wedge   \!\!\!\!\!\!\!\!\!\!\!\! \bigwedge\limits_{X(\bar{w}) \, \in \, \varphi^{+}_i(\bar{X},\bar{y})} \!\!\!\!\!\!\!\!\!\!\!\! \bar{w}\neq\bar{z} \, ) \, \big)\big] \;  \vee \!  \label{eq:part2} \\ 
&\; \ex{\bar{x}'} \ex{\bar{y}}\big( \varphi_i'(\bar{X},\bar{x}',\bar{y})\wedge \bar{x}' < \bar{x}\big) \; \;\; \Big)  \label{eq:part3}
\end{align}
%\martin{agregu\'e par\'entesis en varias f\'ormulas en esta p\'agina y las anteriores}
%\marcelo{OK}

To understand the formula, first notice that the premise of the implication at (\ref{eq:part1}) is true if, and only if, $\varphi_i(\bar{X}, \bar{x})$ is satisfiable. 
Indeed, by Claim~\ref{claim:minusone} we know that if $\ex{\bar{x}} \ex{\bar{y}} \varphi^{\fo}_i(\bar{x},\bar{y})$ is true, then there exist assignments $V$ and $v$ such that $(\A,V,v) \models\varphi_i'(\bar{X},\bar{x},\bar{y})$.
The conclusion of the implication (divided into (\ref{eq:part2}) and (\ref{eq:part3})), takes care that $V(\bar{X}) \neq V^*(\bar{X})$ or $v(\bar{x}) \neq v^*(\bar{x})$.
Here, the first disjunct (\ref{eq:part2}) checks that $V(\bar{X}) \neq V^*(\bar{X})$ by defining that if $\varphi_i'(\bar{X},\bar{x}, \bar{y})$ is satisfied then $V^*(\bar{X}) \subsetneq V(\bar{X})$. 
The second disjunct (\ref{eq:part3}) is satisfied when $v(\bar{x})$ is not the lexicographically smallest tuple that satisfies $\varphi_i$ (i.e. $v(\bar{x}) \neq v^*(\bar{x})$).
Finally, from the previous discussion one can easily check that $\psi_i(\bar{X},\bar{x})$ satisfies the desired property.
